---
title: "Curate the metadata and phylogenetic tree for this project"
author: "Kyle Gontjes"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
require(knitr)
knitr::opts_knit$set(echo = TRUE,warning = FALSE, message = FALSE)  
knitr::opts_knit$set(root.dir='~/Desktop/gl_mount/Analysis/Phyloaware/phylogenetic-resistance-ms/') 
```

# Environment

```{r}
# Packages
packages <- c("readxl","tidyverse","ape","phytools")
lapply(packages,library,character.only=T)

# Functions
source("./scripts/dataset_curation/metadata_curation_functions.R")

# Print environment
Sys.info()
sessionInfo()
```

# Clean the patient metadata

Source: UPENN metadata file (459_patients_data_for_posting.xls)

NFS Path:
/nfs/turbo/umms-esnitkin/Project_Penn_KPC/Analysis/data/459_patients_data_for_posting.xls

## Load and remove isolate with incomplete data

```{r}
# Load in the patient data
metadata_raw <- read_xls('../../data/459_patients_data_for_posting.xls') %>% as.data.frame
paste0("Number of isolates in metadata: ",nrow(metadata_raw))
# Keep only california isolates
metadata <- subset(metadata_raw,state=="CA")
metadata$isolate_no <-  paste0("PCMP_H",metadata$isolate_no)
rownames(metadata) <- metadata$isolate_no 
paste0("Number of isolates in metadata from California: ",nrow(metadata))

# Remove patient with mismatched SENSITIRE isolate number
metadata <- metadata %>% subset(isolate_no != "PCMP_H473")
paste0("Number of isolates in metadata from California with correct SENSITITRE data: ",nrow(metadata))
```

## Update length of stay data

```{r}
# Length of stay
metadata <- metadata %>% mutate(LOSbeforeCx = dplyr::recode(LOSbeforeCx, '-202' = 202))
```

## Recode antibiotic exposure history data

```{r}
## Create new days of therapy (DOT) variables
metadata$AG_day <- metadata %>% select(amikacin_day,gent_day,tobra_day) %>% rowSums()
metadata$third_plus_ceph_day <- metadata %>% select(ceftazidime_day,ctx_day,ceftaroline_day,cefepime_day) %>% rowSums()
metadata$FQ_day <- metadata %>% select(cipro_day,levo_day) %>% rowSums() 
metadata$pseudoPCN_day <- metadata %>% select(zosyn_day,ceftazidime_day) %>% rowSums() 

## Binarize antibiotic exposure history (1 day was chosen)
DOT_vars <- colnames(metadata) %>% subset(grepl("day",.))
for(i in DOT_vars){
  metadata[,paste0(gsub("_day","",i),"_1d")] <-  binarize_abx_history(i,1,metadata)
}

## Binarized data
metadata$any_antibiotic_1d <- ifelse(rowSums(metadata %>% select(any_of(DOT_vars)))>0,1,0)
```

# Antibiotic resistance metadata

Source; Custom SENSITITRE broth microdilution assay

NFS Paths:
/nfs/turbo/umms-esnitkin/Project_Penn_KPC/Analysis/Antibiotic_resistance/data/SENSITITRE
MICs.xlsx

## Load data

```{r}
SENSITITRE <- read_excel("../../Antibiotic_resistance/data/SENSITITRE MICs.xlsx", range = "B1:N460") %>% as.data.frame
SENSITITRE <- SENSITITRE %>% mutate(isolate_no = paste0("PCMP_H",str_remove(gsub('CRKP_','',SENSITITRE$ISOLATES),"^0+"))) %>% `rownames<-`(.$isolate_no)

# PCMP_H472 & PCMP_H473 were excluded due to data reporting issues
SENSITITRE <- subset(SENSITITRE,!isolate_no %in% c("PCMP_H472","PCMP_H473"))
```

## Clean variables

```{r}
## Convert names
SENSITITRE <-  dplyr::rename(SENSITITRE,MVB = "MEV",IR = "IMK", CST = "COL", AMK = "AMI", FOS = "FOS+",PT = "P/T4")

## Fix MVB data due to miscoding (vaborbactom is a constant 8 µg/mL not 4 µg/mL)
table(SENSITITRE$MVB)
SENSITITRE <- SENSITITRE %>% mutate(MVB = dplyr::recode(MVB,"2/4" = "2/8"))

## Create MIC Variable
MIC_variables <- c("Ceftazidime-avibactam"="CZA","Meropenem-vaborbactam"="MVB","Imipenem-relebactam"="IR","Tigeycline"="TGC","Colistin"="CST","Amikacin"="AMK","Plazomicin"="PLZ","MERO","IMI")

## Numerize MIC Data
for(i in MIC_variables){
  SENSITITRE[,paste0(i,"_num")] <- SENSITITRE[,paste0(i)]  %>% gsub("≤","",.) %>% gsub(">","",.) %>% gsub("/.","",.) %>% trimws(.,'both') %>% as.numeric
}

## Factorize Raw MIC Data
for(i in MIC_variables){
  SENSITITRE[,i] <- factorize_MIC(SENSITITRE,i)
}

## Log-2 Transform Data
for(i in MIC_variables){
  SENSITITRE[,paste0(i,"_log_2")] <- log2(SENSITITRE[,paste0(i,"_num")])
}
```

## Convert Raw MIC data to phenotypic resistance categories

Select references:

1\. <https://academic.oup.com/cid/article/72/8/1484/5868454?login=true>

2\.
<https://www.sciencedirect.com/science/article/pii/S0255085721041062>

```{r}
SENSITITRE <- SENSITITRE %>% mutate(
  ### Meropenem (MERO): ≤1/2/≥4
  MERO_cat = ifelse(MERO_num <= 1,"S",ifelse(MERO_num >= 4,"R","I")),
  ### Imipenem (IMI): ≤1/2/≥4
  IMI_cat = ifelse(IMI_num <= 1,"S",ifelse(IMI_num >= 4,"R","I")),
  ### Ceftazidime-avibactam: Susceptible (MIC≤8/4), Resistant (MIC≥16/4)  
  CZA_cat = ifelse(CZA_num <= 8,"S",ifelse(CZA_num >= 16,"R","")),
  ### Meropenem-vaborbactam (MVB): Susceptible (MIC≤4/8), Intermediate (MIC=8/8), Resistant (MIC≥16/8)  
  MVB_cat = ifelse(MVB_num <= 4,"S",ifelse(MVB_num == 8,"I","R")),
  ### Imipenem-relebactam (I-R): Susceptible (MIC≤1/4), Intermediate (MIC=2/4), Resistant (MIC≥4/4)  
  IR_cat = ifelse(IR_num <= 1,"S",ifelse(IR_num == 2,"I","R")),
  ### Tigecycline (TGC): Susceptible (MIC≤2), Intermediate (MIC=4), Resistant (MIC≥8)  
  TGC_cat = ifelse(TGC_num <= 2,"S",ifelse(TGC_num == 4,"I","R")),
  ### Colistin (CST): Intermediate (MIC≤2), Resistant (MIC≥4)
  CST_cat = ifelse(CST_num <= 2,"I",ifelse(CST_num >= 4,"R","")),
  ### Amikacin (AMK): Susceptible (MIC≤16), Intermediate (MIC=32), Resistant (MIC≥64)   
  AMK_cat =  ifelse(AMK_num <= 16,"S",ifelse(AMK_num == 32,"I","R")),
  ### Plazomicin (PLZ): Susceptible (MIC≤2), Intermediate (MIC=4), Resistant (MIC≥8) 
  PLZ_cat = ifelse(PLZ_num <= 2,"S",ifelse(PLZ_num == 4,"I","R")))

## Factorize Categorical Data
MIC_variables_cat <- paste0(MIC_variables,"_cat") 
for(i in MIC_variables_cat){
  SENSITITRE[,i] <- factor(as.vector(SENSITITRE[,i]),
                           levels = c("S","I","R"),
                           labels = c("Susceptible","Intermediate","Resistant"))
}

## Create Binary Non-Susceptible Variable
MIC_variables_no_CST <- c("MERO","IMI","MVB","IR","CZA","AMK","PLZ","TGC")
for(i in MIC_variables_no_CST){
  SENSITITRE[,paste0(i,"_dich")] <- binarize_MIC(SENSITITRE,paste0(i,"_cat"))
}
SENSITITRE$CST_dich <- ifelse(SENSITITRE$CST_cat == "Resistant","Non-Susceptible","Susceptible")

## Factorize Dichotomous Data
MIC_variables_dich <- paste0(MIC_variables,"_dich") 
for(i in MIC_variables_dich){
  SENSITITRE[,i] <- factor(as.vector(SENSITITRE[,i]),
                           levels = c("Susceptible","Non-Susceptible"),
                           labels = c("Susceptible","Non-Susceptible"))
}

## Numerical Dichotemous Variable for Resistance
for(i in MIC_variables_dich){
  SENSITITRE[,paste0(i,"_num")] <- ifelse(SENSITITRE[,i] == "Susceptible",0,1)
}
```

### Simplify BL/BLI Non-susceptibility

```{r}
# BL/BLI Resistance
SENSITITRE$blbli_dich <- ifelse(SENSITITRE$MVB_dich_num==1 | SENSITITRE$IR_dich_num==1,"Non-Susceptible","Susceptible")
SENSITITRE$blbli_dich_num <- ifelse(SENSITITRE$MVB_dich_num==1 | SENSITITRE$IR_dich_num==1,1,0)
```

## Curate lab-abstracted phenotypes

```{r}
# Gentamicin Recode
metadata$gentamicin_mic <- metadata$gent_mic %>% {ifelse(.==">-16",">=16",.)} %>% gsub(">=","≥",.) %>% gsub("<=","≤",.)
metadata$gentamicin_int <- ifelse(metadata$gentamicin_mic %in% c("≤1","2","4"),"S",ifelse(metadata$gentamicin_mic =="8","I","R") )
   
# Bactrim Recode
metadata$bactrim_mic <- metadata$bactrim_mic %>% {ifelse(.=="4","40",.)} %>% gsub(">=","≥",.) %>% gsub("<=","≤",.)
metadata$bactrim_int <- metadata$bactrim_mic %>% {ifelse(.=="40","S",metadata$bactrim_int)}
metadata$bactrim_int <- metadata$bactrim_mic %>% {ifelse(.=="40","S", ifelse(.=="≥320","R",metadata$bactrim_int))}

MIC_variables <- c("bactrim_mic","gentamicin_mic") 

factorize_MIC_lab <- function(dataset,resistance){
  values <- as.vector(unique(dataset[,resistance]))
  combination <- ifelse(length(grep("/",values))>0,"yes","no")
  if(combination == "no"){
    level <- c(
      #Endpoint 
      values[grep("≤|<|<=",values)], 
      #if non endpoint included 
      #non-endpoint sorted
      as.character(sort(as.numeric(values[!grepl(">|>=|≥|≤|<|<=", values)]))),
      #Endpoint
      values[grep(">|>=|≥",values)])
  }
  if(combination == "yes"){
    values_no_in <- gsub("/.*","",values)
    in_value <- gsub(".*\\/","",values)
    level <- c(
      #Endpoint 
      values_no_in[grep("≤",values_no_in)], 
      #if non endpoint included 
      #non-endpoint sorted
        as.character(sort(as.numeric(values_no_in[!grepl("≤|>", values_no_in)]))),
      #Endpoint
      values_no_in[grep(">",values_no_in)])
    level <- paste0(level,"/",in_value)
  }    
  factorized <- factor(as.vector(dataset[,resistance]),levels = level)
  return(factorized)
}

for(i in MIC_variables){
  metadata[,i] <- factorize_MIC_lab(metadata,i)
}

## Factorize Categorical Data
MIC_variables <- c("bactrim","gentamicin")
for(i in MIC_variables){
  metadata[,paste0(i,"_cat")] <- factor(as.vector(metadata[,paste0(i,"_int")]),
                           levels = c("S","I","R"),
                           labels = c("Susceptible","Intermediate","Resistant"))
}

## add bactrim (TMP_SMX), gentamicin
metadata$TMP_SMX_cat <- metadata$bactrim_cat
metadata$TMP_SMX_mic <- metadata$bactrim_mic
metadata$TMP_SMX_dich <- ifelse(metadata$bactrim_cat == "Resistant","Non-Susceptible","Susceptible")
metadata$TMP_SMX_dich_num <- ifelse(metadata$bactrim_cat == "Resistant",1,0)

## Gentamicin 
metadata$gentamicin_dich <- ifelse(metadata$gentamicin_cat == "Susceptible","Susceptible","Non-Susceptible")
metadata$gentamicin_dich_num <- ifelse(metadata$gentamicin_cat == "Susceptible",0,1)
```

## Curate sequence type data

Source: Kleborate version 3.1.3

NFS Paths:
/nfs/turbo/umms-esnitkin/Project_Penn_KPC/Analysis/Phyloaware/phylogenetic-resistance-ms/data/kleborate/klebsiella_pneumo_complex_output.txt

```{r}
# Load data
kleborate <- read.delim(file='./data/kleborate/klebsiella_pneumo_complex_output.txt',sep='\t')

# Get isolate name from assembly name
kleborate<- kleborate %>% mutate(isolate_no = gsub("_contigs_l1000","",strain)) %>% `rownames<-`(.$isolate_no) 

# Clean ST Variable
kleborate <- kleborate %>% mutate(ST = gsub("ST","",ST) %>% gsub("-1LV","*",.),
                                  ST258 = ifelse(ST == "258",1,0))
```

# Create metadata dataframe

1.  Drop to only ST258 genomes
2.  Drop to only isolates from california

```{r}
# Merge metadata and sensititre data
df <- left_join(metadata,SENSITITRE,by="isolate_no") 
# Only keep ST258 isolates
df <- left_join(df,kleborate %>% select(isolate_no,ST,ST258))
df <- subset(df,ST258==1)
paste0("ST258 isolates from California: ",nrow(df))
```

# Phylogenetic tree

```{r}
tr <- read.tree(file="../../../Sequence_data/variant_calling/2025-02-17_snpkit_Penn_KPC_ST258/output_files/2025_02_23_21_11_49_core_results/phylokit/results/IQtree/Penn_KPC_ST258.treefile")  %>% 
                    drop.tip(
                    # Reference Sequence: genome w/o sensitire) & PCMP_H243 (outgroup)
                    c('gi|661922017|gb|CP008827.1|','PCMP_H243','PCMP_H473'))

# California only isolates
tr_ca <- keep.tip(tr,df %>% subset(state=='CA') %>% .$isolate)
# Midpoint rooted tree 
tr_ca_midpoint = phytools::midpoint.root(tr_ca) 
paste0("ST258 isolates from California with assembly: ",length(tr_ca$tip.label))

# Save  
write.tree(tr_ca_midpoint,"./data/tree/tree.treefile")
```

# Only keep isolates with entry in phylogenetic tree (i.e., good genome

```{r}
tr <- keep.tip(tr,df$isolate_no)
df <- df %>%.[match(as.vector(tr$tip.label), .$isolate_no),]
rownames(df) <- df$isolate_no
```

# Get clade data

```{r}
library(ggtree)
root_number <-  tr_ca_midpoint$edge[,1] %>% as.numeric %>% min
clade_nodes <-  tr_ca_midpoint$edge[tr_ca_midpoint$edge[,1] == root_number,] %>% .[,2]  
 
ggtree(tr_ca_midpoint) + geom_point2(aes(subset=node==clade_nodes[1]), color='red', size=5)  + geom_point2(aes(subset=node==clade_nodes[2]), color='blue', size=5)
 
clade_1_descendants = tr_ca_midpoint$tip.label[ phangorn::Descendants(x = tr_ca_midpoint,node = clade_nodes[[1]],type='tips') %>% unlist]
clade_2_descendants = tr_ca_midpoint$tip.label[ phangorn::Descendants(x = tr_ca_midpoint,node = clade_nodes[[2]],type='tips') %>% unlist]

df$clade_I <- ifelse(df$isolate_no %in% clade_1_descendants,"Clade I","Clade II")
gheatmap(ggtree(tr_ca_midpoint),df %>% select(clade_I))
```

# Save the metadata file to dataset folder

```{r}
saveRDS(df,"./data/dataset/df.RDS") 
```
